---
title: "BEACON: Correcting Surveillance Undercount for AI/AN Gestational Diabetes in Oklahoma"
author: "Emily N. Peterson"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## Background

### The Problem

American Indian and Alaska Native (AI/AN) women face some of the highest rates
of gestational diabetes mellitus (GDM) in the United States — roughly twice
the rate of non-Hispanic White women. Yet state-level surveillance data are
known to undercount AI/AN births and diagnoses, both because of racial
misclassification and because a substantial share of Cherokee Nation members
receive prenatal care exclusively through Cherokee Nation Health Services
rather than providers who report to the Oklahoma State Department of Health
(OSDH).

The result is a systematic downward bias: the OSDH-reported AI/AN GDM rate in
a county reflects not just the true disease burden but also how completely the
surveillance system detects it. Counties with strong Cherokee Nation clinical
infrastructure may look statistically *similar* to counties with very low true
prevalence, simply because their cases flow through a parallel health system.

### Cherokee Nation as an Anchor

Cherokee Nation Public Health (CNPH) maintains its own clinical registry
covering GDM diagnoses for births at CN health facilities. In the 7 Oklahoma
counties that have CN service coverage, this registry provides a near-complete
count of GDM among Cherokee mothers — a gold-standard anchor that the state
system cannot provide on its own.

BEACON (**B**ayesian **E**stimation and **C**orrection for **A**I/AN **O**utcomes
using **N**ative-data) jointly models the OSDH and CN data streams to recover
the latent true AI/AN GDM prevalence surface across all 77 Oklahoma counties.

---

## The BEACON Model

### Three Latent County-Level Surfaces

Let $c = 1, \ldots, 77$ index Oklahoma counties. The model estimates three
latent quantities simultaneously:

| Symbol | Name | Interpretation |
|--------|------|----------------|
| $\rho_c$ | True AI/AN GDM prevalence | Probability a randomly selected AI/AN birth in county $c$ results in a GDM diagnosis |
| $\varphi_c$ | OSDH sensitivity | Probability OSDH captures a true case in county $c$ |
| $\kappa_c$ | CN detection probability | Probability CN health system captures a Cherokee birth in county $c$ |

Each surface lives on $(0,1)$ and is modeled on the logit scale with
covariate terms and spatial random effects.

### Likelihood

**OSDH data** (Eq. 1). For county $c$ with AI/AN births $n_c^{\text{OSDH}}$,
the observed OSDH GDM count $y_c^{\text{OSDH}}$ is:
$$
y_c^{\text{OSDH}} \mid n_c^{\text{OSDH}} \sim \text{Binomial}\!\left(n_c^{\text{OSDH}},\; \rho_c \cdot \varphi_c\right)
$$

OSDH only observes the *attenuated* rate $\rho_c \cdot \varphi_c$. Counties
suppressed by OSDH (fewer than 5 AI/AN births) contribute no direct likelihood.

**Cherokee Nation data** (Eq. 2). For the $k = 1, \ldots, 7$ CN-covered
counties with CN births $n_k^{\text{CN}}$:
$$
y_k^{\text{CN}} \mid n_k^{\text{CN}} \sim \text{Binomial}\!\left(n_k^{\text{CN}},\; \rho_k \cdot \kappa_k\right)
$$

The shared parameter $\rho_c$ appears in both likelihoods — this is the
identification mechanism. Because each data source attenuates $\rho_c$ by a
*different* nuisance parameter ($\varphi_c$ vs. $\kappa_c$), their joint
information can separate the true signal from the detection bias.

### Regression Models

**True prevalence** (Eq. 3):
$$
\text{logit}(\rho_c) = \beta_0 + \beta_1 \text{SNAP}_c + \beta_2 \text{LowAccess}_c + \beta_3 \text{AdequateCare}_c + v_c^{(p)} + u_c^{(p)}
$$

Covariates: proportion of county receiving SNAP ($+$), proportion with low
food access ($+$), proportion of births with adequate prenatal care ($-$).

**OSDH sensitivity** (Eq. 4):
$$
\text{logit}(\varphi_c) = \gamma_0 + \gamma_1 \log(\hat{p}_c^{\text{AIAN}}) + v_c^{(\varphi)} + u_c^{(\varphi)}
$$

Sensitivity is modeled as a function of the log-proportion AI/AN in the
county. Counties with a larger AI/AN share are expected to have better-tuned
surveillance systems.

**CN detection probability** (Eq. 5):
$$
\text{logit}(\kappa_c) = \alpha_0 + \alpha_1 \cdot \text{ClinicCount}_c + \alpha_3 \cdot \text{Spillover}_c
$$

CN detection has no spatial random effect. The two covariates capture direct
CN clinical presence (number of CN facilities in county $c$) and a
distance-decay spillover term measuring proximity to CN facilities in
neighboring counties:
$$
\text{Spillover}_c = \sum_{j \in \mathcal{J}} \exp\!\left(-d_{cj}/75\right)
$$
where $d_{cj}$ is the distance in km from county $c$'s centroid to CN clinic
$j$, and the sum excludes clinics in county $c$'s own territory (already
captured by ClinicCount).

The 75-km bandwidth reflects the practical travel range for prenatal
appointments — enough to reach an adjacent county, but insufficient to span
the state.

### Random Effects

Both $\rho_c$ and $\varphi_c$ include a BYM2-style decomposition into an ICAR
structured component $v_c$ (spatial autocorrelation) and an unstructured iid
component $u_c$:
$$
v_c^{(\cdot)} \sim \text{ICAR}(\tau_{v}^{(\cdot)}), \quad u_c^{(\cdot)} \sim \mathcal{N}(0, \sigma_{u}^{(\cdot)2})
$$

The $\kappa_c$ surface has no random effect: with GDM data from only 7 CN
counties, a spatial random effect would over-smooth CN-county estimates into
adjacent non-CN counties.

### Priors

All regression coefficients receive weakly informative Normal priors centered
near zero; $\beta_0$ is centered at $\text{logit}(0.12) \approx -2$
(roughly 10–15% GDM prevalence) and $\alpha_0$ at $-1$ (roughly 27% baseline
CN detection, recognizing incomplete capture even within CN territory). Variance
components receive half-Normal(0, 2.5) priors.

---

## Synthetic Data Demonstration

Because the Cherokee Nation clinical data are governed by a Tribal data use
agreement, the demo below uses **synthetic data** generated with parameters
fixed at the posterior means from the fitted model. The synthetic data are
designed to mirror the structure of the real analysis: 77 counties, realistic
covariate distributions, suppression of small AI/AN counts, and CN clinical
data for 7 service counties.

> **Note:** Run `demo_aian_gdm_oklahoma.R` (or `make_example_outputs.R` for a
> faster version) before rendering this vignette. Outputs are loaded from
> `output/`.

### Setup

```{r setup}
library(tidyverse)
library(tigris)
library(sf)
library(spdep)

options(tigris_use_cache = TRUE)
```

### Oklahoma County Spatial Structure

```{r spatial}
OKcounties <- tigris::counties("Oklahoma", cb = FALSE)
OKcounties <- OKcounties[order(OKcounties$NAME), ]
county_names <- OKcounties$NAME
C            <- length(county_names)

ok_map <- tigris::counties(state = "OK", cb = TRUE, class = "sf")
ok_map <- ok_map[order(ok_map$NAME), ]

cat(sprintf("Counties: %d\n", C))
```

### Synthetic Covariates

The demo script generates three types of covariates:

- **$x_c$** — true prevalence model: SNAP receipt (`propSNAP`), low food access
  (`prop_lowaccess`), adequate prenatal care (`adequate_care`)
- **$w_c$** — sensitivity model: log-proportion AI/AN (`prop_aian`)
- **$d_c$** — CN detection: clinic count (`n_clinics`) + distance-decay
  spillover (`cn_spillover`)

```{r covariates}
set.seed(2025)

propSNAP       <- rbeta(C, 2,  8)
prop_lowaccess <- rbeta(C, 2, 10)
adequate_care  <- rbeta(C, 7,  2)
prop_aian      <- pmax(rbeta(C, 1, 18), 0.001)

cn_county_names <- c("Adair", "Cherokee", "Delaware", "Mayes",
                     "Muskogee", "Rogers", "Sequoyah")
cn_idx    <- which(county_names %in% cn_county_names)
n_clinics <- rep(0L, C)
n_clinics[cn_idx] <- sample(1:2, length(cn_idx), replace = TRUE)

# Distance-decay spillover from publicly known CN facility locations
cn_clinic_coords <- data.frame(
  lat    = c(35.9106, 35.9098, 35.7297, 36.4212, 36.7480,
             35.4587, 35.8144, 36.2929, 36.7006, 36.6408),
  long   = c(-94.9710, -94.9478, -95.3104, -94.8043, -95.9724,
             -94.7879, -94.6286, -95.1533, -95.6380, -95.1564),
  county = c("Cherokee", "Cherokee", "Muskogee", "Delaware", "Washington",
             "Sequoyah", "Adair", "Mayes", "Nowata", "Craig")
)
lambda_km           <- 75
cn_clinics_sf       <- st_as_sf(cn_clinic_coords, coords = c("long", "lat"), crs = 4326) %>%
  st_transform(crs = 5070)
county_centroids_sf <- ok_map %>% st_transform(crs = 5070) %>%
  st_centroid() %>% arrange(NAME)
dist_km             <- matrix(as.numeric(st_distance(county_centroids_sf, cn_clinics_sf)),
                              nrow = C) / 1000
same_county_mask    <- outer(county_centroids_sf$NAME, cn_clinic_coords$county, FUN = "==")
decay_matrix        <- exp(-dist_km / lambda_km)
decay_matrix[same_county_mask] <- 0
cn_spillover        <- rowSums(decay_matrix)
```

The spillover surface decays steeply with distance. @fig-spillover shows that
only counties immediately adjacent to the CN core service area receive
meaningful spillover weight.

```{r fig-spillover, fig.cap="Distance-decay CN spillover covariate. Darker counties are closer to at least one CN clinical facility."}
ok_map %>%
  left_join(data.frame(NAME = county_names, spillover = cn_spillover),
            by = "NAME") %>%
  ggplot() +
  geom_sf(aes(fill = spillover), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       name = "Spillover\nweight") +
  theme_minimal(base_size = 11) +
  labs(title    = "CN Distance-Decay Spillover Covariate",
       subtitle = expression(Sigma[j] ~ exp(-d[cj]/75 ~ km) ~ ", own-county excluded"))
```

### True Parameter Values

The synthetic data use these parameters, fixed at posterior means from the
fitted model. The model's job is to recover them.

```{r true-params}
beta0  <- -1.56;  beta1  <-  3.50;  beta2  <- -0.36;  beta3  <- -0.36
gamma0 <- -0.46;  gamma1 <- -0.36
alpha0 <- -1.48;  alpha1 <-  0.97;  alpha3 <-  1.00

sigma_p_v   <- 0.345;  sigma_p_u   <- 0.321
sigma_phi_v <- 0.540;  sigma_phi_u <- 0.288

tibble(
  Parameter = c("β₀ (GDM intercept)", "β₁ (SNAP)", "β₂ (Low food access)",
                "β₃ (Adequate care)", "γ₀ (Sensitivity intercept)",
                "γ₁ (log prop. AIAN)", "α₀ (CN coverage intercept)",
                "α₁ (CN clinic count)", "α₃ (Spillover)",
                "σ_p,v", "σ_p,u", "σ_φ,v", "σ_φ,u"),
  `True value` = c(beta0, beta1, beta2, beta3, gamma0, gamma1,
                   alpha0, alpha1, alpha3,
                   sigma_p_v, sigma_p_u, sigma_phi_v, sigma_phi_u)
) |> knitr::kable(digits = 3)
```

Key interpretations on the probability scale:

- $\text{logit}^{-1}(\beta_0) \approx$ `r round(plogis(beta0), 3)` — baseline GDM
  prevalence in a county at mean covariates, no spatial effect
- $\text{logit}^{-1}(\gamma_0) \approx$ `r round(plogis(gamma0), 3)` — baseline OSDH
  sensitivity (roughly 39% of cases are captured at the average AI/AN share)
- $\text{logit}^{-1}(\alpha_0) \approx$ `r round(plogis(alpha0), 3)` — baseline CN
  detection in a county with no CN facilities and zero spillover

---

## NIMBLE Model Code

The full NIMBLE model definition is reproduced here for reference. See
`demo_aian_gdm_oklahoma.R` for a directly runnable version.

```{r nimble-model, eval=FALSE}
library(nimble)

myCode <- nimbleCode({

  # OSDH likelihood [Eq. 1]
  for (i in 1:N_OSDH) {
    y_OSDH_i[i] ~ dbin(rho_error_c[getc_i[i]], n_OSDH_i[i])
  }

  # Cherokee Nation likelihood [Eq. 2]
  for (k in 1:N_CN) {
    y_CN_k[k] ~ dbin(rho_true_c[getc_k[k]] * kappa_c[getc_k[k]], n_CN_k[k])
  }

  for (c in 1:Ncounties) {

    rho_error_c[c] <- rho_true_c[c] * phi_c[c]   # attenuated OSDH rate

    # True prevalence [Eq. 3]
    logit(rho_true_c[c]) <- beta0 +
                            beta1 * propSNAP[c] +
                            beta2 * prop_lowaccess[c] +
                            beta3 * adequate_care[c] +
                            v_p_c[c] + u_p_c[c]

    # OSDH sensitivity [Eq. 4]
    logit(phi_c[c]) <- gamma0 +
                       gamma1 * log(prop_aian[c]) +
                       v_phi_c[c] + u_phi_c[c]

    # CN detection [Eq. 5] — no spatial random effect
    logit(kappa_c[c]) <- alpha0 +
                         alpha1 * n_clinics[c] +
                         alpha3 * cn_spillover[c]

    u_p_c[c]   ~ dnorm(0, sd = sigma_p_u)
    u_phi_c[c] ~ dnorm(0, sd = sigma_phi_u)
  }

  # ICAR spatial effects
  v_p_c[1:Ncounties]   ~ dcar_normal(adj[1:L], weights[1:L], num[1:Ncounties],
                                     tau = 1 / (sigma_p_v^2),   zero_mean = 1)
  v_phi_c[1:Ncounties] ~ dcar_normal(adj[1:L], weights[1:L], num[1:Ncounties],
                                     tau = 1 / (sigma_phi_v^2), zero_mean = 1)

  # Priors
  beta0 ~ dnorm(-2,  sd = 1)
  beta1 ~ dnorm(0.5, sd = 5);  beta2 ~ dnorm(0.5, sd = 5);  beta3 ~ dnorm(0.5, sd = 5)
  gamma0 ~ dnorm(0,  sd = 1);  gamma1 ~ dnorm(0, sd = 1)
  alpha0 ~ dnorm(-1, sd = 1);  alpha1 ~ dnorm(0, sd = 1);  alpha3 ~ dnorm(0, sd = 1)

  sigma_p_u   ~ T(dnorm(0, sd = 2.5), 0.0001, )
  sigma_p_v   ~ T(dnorm(0, sd = 2.5), 0.0001, )
  sigma_phi_u ~ T(dnorm(0, sd = 2.5), 0.0001, )
  sigma_phi_v ~ T(dnorm(0, sd = 2.5), 0.0001, )
})
```

---

## Results

The following sections load pre-computed posterior summaries from
`output/demo_results_table.csv` (generated by running
`demo_aian_gdm_oklahoma.R`). County-level posteriors are loaded from
`output/demo_county_posteriors.csv`.

```{r load-results}
results_file  <- "output/demo_results_table.csv"
counties_file <- "output/demo_county_posteriors.csv"

if (!file.exists(results_file)) {
  stop("Run demo_aian_gdm_oklahoma.R (or make_example_outputs.R) first to generate output/.")
}

results_table   <- read.csv(results_file)
county_post     <- read.csv(counties_file)
```

### Scalar Parameter Recovery

@tbl-recovery shows posterior means, 95% credible intervals, and
Gelman–Rubin $\hat{R}$ for each scalar parameter alongside the true synthetic
value.

```{r tbl-recovery}
#| label: tbl-recovery
#| tbl-cap: "Posterior summaries and convergence diagnostics. True values are the fixed parameters used to generate the synthetic data. Rhat < 1.1 indicates adequate chain mixing."
results_table |>
  mutate(
    `95% CrI` = sprintf("[%.3f, %.3f]", LCI_95, UCI_95),
    Rhat      = round(Rhat, 3),
    ESS       = round(ESS, 0)
  ) |>
  select(Parameter, True_Value, Post_Mean, Post_SD, `95% CrI`, Rhat, ESS) |>
  knitr::kable(digits = 3)
```

A well-fitting run will show posterior means close to the true values with
$\hat{R} \leq 1.10$ for all parameters. The 95% credible interval should
contain the true value for roughly 95% of parameters across repeated synthetic
datasets.

### County-Level Maps

```{r maps-setup}
map_df <- ok_map %>%
  left_join(county_post, by = c("NAME" = "county"))
```

#### Adjusted AI/AN GDM Prevalence

@fig-rho shows the posterior mean of $\rho_c^{\text{true}}$ — the adjusted
prevalence after accounting for OSDH sensitivity. Compare this to the naive
observed OSDH rate (@fig-obs): counties that are suppressed or poorly detected
by OSDH receive information through the spatial model and the CN anchor.

```{r fig-rho, fig.cap="Posterior mean adjusted AI/AN GDM prevalence. Grey counties had OSDH data suppressed (< 5 AI/AN births) and are estimated entirely through the spatial model and CN likelihood."}
ggplot(map_df) +
  geom_sf(aes(fill = rho_mean), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "Purples", direction = 1,
                       name = expression(hat(rho)[c]^{true}),
                       limits = c(0, 0.45), oob = scales::squish,
                       na.value = "grey80") +
  theme_minimal(base_size = 11) +
  labs(title    = "Posterior Mean Adjusted AI/AN GDM Prevalence",
       subtitle = "Synthetic data demonstration")
```

```{r fig-obs, fig.cap="Naive observed OSDH rate (GDM cases / AI/AN births). Grey counties are suppressed. The BEACON model corrects both suppressed counties and systematic underdetection in reported counties."}
ggplot(map_df) +
  geom_sf(aes(fill = obs_rate), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "Purples", direction = 1,
                       name = "Observed\nrate",
                       limits = c(0, 0.45), oob = scales::squish,
                       na.value = "grey80") +
  theme_minimal(base_size = 11) +
  labs(title    = "Naive OSDH-Observed AI/AN GDM Rate (unadjusted)",
       subtitle = "Synthetic data demonstration — grey = suppressed")
```

#### OSDH Sensitivity Surface

@fig-phi maps $\hat{\varphi}_c$ — the estimated probability that a true AI/AN
GDM case in county $c$ appears in OSDH records. High-AIAN-share counties
(eastern Oklahoma) tend to have higher sensitivity because providers in those
communities are more experienced at capturing AI/AN race/ethnicity correctly.

```{r fig-phi, fig.cap="Posterior mean OSDH sensitivity. Values near 1 indicate high data quality; values near 0 indicate substantial undercounting."}
ggplot(map_df) +
  geom_sf(aes(fill = phi_mean), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "RdYlGn", direction = 1,
                       name = expression(hat(phi)[c]),
                       limits = c(0, 1)) +
  theme_minimal(base_size = 11) +
  labs(title    = "Posterior Mean OSDH Sensitivity",
       subtitle = "Synthetic data demonstration")
```

#### CN Detection Probability

@fig-kappa shows $\hat{\kappa}_c$ for all 77 counties. Only the 7 CN-service
counties contribute direct likelihood information for $\kappa_c$; for all other
counties $\kappa_c$ is determined entirely by the distance-decay spillover
covariate and the $\alpha_0$ intercept.

```{r fig-kappa, fig.cap="Posterior mean CN detection probability. The 7 CN-service counties (darker) are estimated primarily from the Cherokee Nation clinical data; the remaining counties reflect the distance-decay covariate only."}
cn_map <- map_df %>% filter(NAME %in% cn_county_names)

ggplot() +
  geom_sf(data = map_df,  fill = "grey92", color = "white", linewidth = 0.2) +
  geom_sf(data = map_df,  aes(fill = kappa_mean), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       name = expression(hat(kappa)[c]),
                       limits = c(0, 1), na.value = "grey92") +
  theme_minimal(base_size = 11) +
  labs(title    = "Posterior Mean CN Detection Probability",
       subtitle = "Synthetic data demonstration")
```

---

## Summary

BEACON separates the signal from the noise by jointly modeling two
complementary surveillance streams:

1. **OSDH data** — broad geographic coverage, subject to racial
   misclassification and incomplete provider reporting ($\varphi_c < 1$)
2. **CN clinical data** — complete within Cherokee Nation service counties,
   but only observed for women who use CN facilities ($\kappa_c < 1$)

The shared true prevalence $\rho_c$ is identified because the two sources
attenuate it by different, covariate-driven mechanisms. Spatial borrowing
through ICAR random effects smooths estimates for suppressed counties and
stabilizes inference in counties with very few AI/AN births.

The model is implemented in NIMBLE and can be run end-to-end using
`demo_aian_gdm_oklahoma.R`.

---

## Session Info

```{r session-info}
sessionInfo()
```
